\speculake: Hiding Malware in Speculative Execution

### 2018-12-13 paper/architecture.tex
Updated figure 3 and speculative limitations subsection

\subsubsection{Cache Miss Duration}
When executing instructions speculatively we rely on a memory load of a function
pointer from uncached memory. Thus, one potential limit on our computation comes
in the form of the time it takes for the memory read to return with a result
(and for the CPU to determine the result was mis-predicted).
We measured the number of cycles a
cache miss takes to return by artificially evicting an item from cache and
timing reads from its address.
Figure~\ref{fig:cache-miss} shows the CDF of cycles taken. In the typical case,
an evicted item takes approximately 300 clock cycles to load from the Level 3 cache (L3), which
would allow a limit of roughly 200 speculative instructions to be
executed during that time. We note that when an item is not in L3, it takes
considerably longer to load, in theory allowing for thousands of speculative
instructions in a significant fraction of runs.

@@

We also tested whether instruction complexity or data dependencies impact the
number of instructions that can be completed. We find that data dependencies and
instruction complexity both have an impact on the number of instructions that
can be executed. Instruction complexity is determined by the number of $\mu$~ops
that the instruction uses, which appears to be what is tracked in the ROB. For
instance, on our Skylake architecture, the 64-bit \texttt{idiv} instruction
takes 57~$\mu$~ops, and we can execute up to 3 of them in the speculative world.
Meanwhile, we can execute up to 12 32-bit \texttt{idiv} instructions, which each
take 10~$\mu$~ops~\cite{intel-instruction-tables}. This suggest we can execute
on the order of 175~$\mu$~ops before the speculative world expires.

@@

We find that when executing speculatively, the number of instructions completed
is rarely limited by cache miss duration. While data dependency and ROB size
have an effect on the number of instructions that can be executed speculatively
Figure~\ref{fig:spec-capacity-np} demonstrates that the number of $\mu$-ops of
the instructions is the major limiting factor that define an upper boundary of
approximately 150 instructions\footnote{While \texttt{nop} is able to execute up
to the full 220 ROB capacity, instructions that do useful work (and/or use
multiple $\mu$ ops) cannot reach this limit}.

>>>>

\subsubsection{Cache Miss Duration}
When executing instructions speculatively we rely on a memory load of a function
pointer from uncached memory. Thus, one limit on our computation comes
in the form of the time it takes for the memory read to return with a result
(and for the CPU to determine the result was mis-predicted).
We measured the number of cycles a
cache miss takes to return by artificially evicting an item from cache and
timing reads from its address.
Figure~\ref{fig:cache-miss} shows the CDF of cycles taken. In the typical case,
an evicted item takes approximately 200 clock cycles to load from the Level 3 cache (L3), which
would allow a limit of roughly 200 speculative instructions depending on
specific cycles per instruction (CPI), to be
executed during that time. We note that when an item is not in L3, it takes
considerably longer to load, in theory allowing for thousands of speculative
instructions in a significant fraction of runs.

@@

We also tested whether instruction complexity or data dependencies impact the
number of instructions that can be completed. We find that data dependencies and
instruction complexity both have an impact on the number of instructions that
can be executed. Instruction complexity is determined by the number of $\mu$~ops
that the instruction uses, which appears to be what is tracked in the ROB. For
instance, on our Skylake architecture, the 64-bit \texttt{idiv} instruction
takes 57~$\mu$~ops, and we can execute up to 3 of them in the speculative world.
Meanwhile, we can execute up to 18 32-bit \texttt{idiv} instructions, which each
take 10~$\mu$~ops~\cite{intel-instruction-tables}. This suggest we can execute
on the order of 175~$\mu$~ops before the speculative world expires.

@@

We find that when executing speculatively, the number of instructions completed
has a soft limit and a hard limit. The duration in cycles applies a soft limit, 
as shown in Figure~\ref{fig:spec-capacity-np} with the \texttt{idiv (r32)},
\texttt{mul}, and \texttt{aesdec} instructions which see a steep drop in in hit
fraction, but continue to have a non-zero signal.  With all 
instructions a hard limit is imposed by the number of CPU micro-operations ($\mu$-ops)
that compose the individual instruction. Figure~\ref{fig:spec-capacity-np} demonstrates that the number of $\mu$-ops of
the instructions is the major limiting factor that define an upper boundary of
approximately 150 instructions\footnote{While \texttt{nop} is able to execute up
to the full 220 ROB capacity, instructions that do useful work (and/or use
multiple $\mu$-ops) cannot reach this limit. On top of this data dependency and 
execution unit availability add noise further complicating the exact number of instructions
that can be executed speculatively.}

### 2018-12-13 paper/figures.tex
Updated figure 3 and speculative limitations subsection

\newcommand{\FigSpecMeasureNP}{
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/Speculative_measurements_no-parity.png}
    \caption{\textbf{Speculative limits}\,---\,%
    To find the number of instructions we can execute speculatively we placed a signal gadget
    after an increasing number of instructions and measured the fraction of times the 
    loaded value was subsequently in cache. However more complex instructions have stricter 
    constraints applied at two limits that we have identified, cache latency cycle limitation, and 
    the CPU's reorder buffer~(ROB) size as measured by instruction $\mu$-ops. We find that the
    hit fraction for instructions with a ratio of cycle latency to $\mu$~ops greater than 1 undergo a
    steep drop to \~0.1 hit fraction before falling definitively to 0. The drop in hit fraction 
    indicates that the duration of the test in cycles is beyond the fist step in cache latency as shown in 
    Figure~\ref{fig:cache-miss} (approx. 200 cycles). If the instruction sequence does not reach the 
    cycle count limit we see only a sharp drop in hit fraction to 0 indicating the limiting factor 
    is ROB size.  For example, the \texttt{idiv (r32)} instruction we identify 18 as the maximum number of 
    instruction limited by ROB (with ~15 instructions for signal gadget), with a soft limit of 
    12 instructions enforced by the cache latency cycle limitation. 
    We measure the ROB size by timing two cache accesses with a variable number of 
    instructions between them and note that this aligns with the measured maximum \texttt{nop} instrustions.
    }
    \label{fig:spec-capacity-np}
\end{figure*}
}

>>>>

\newcommand{\FigSpecMeasureNP}{
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/Speculative_measurements_no-parity.png}
    \caption{\textbf{Speculative limits}\,---\,%
    To find the number of instructions we can execute speculatively we placed a signal gadget
    after an increasing number of instructions and measured the fraction of times the 
    loaded value was subsequently in cache. We identify two limitation on the number of instructions
    accomplishable speculatively -- cache latency, and the CPU's reorder buffer~(ROB) size. We find that the
    hit fraction for instructions with a ratio of cycle latency to $\mu$~ops greater than 1 undergo a
    steep drop to \~0.1 hit fraction before falling definitively to 0. The drop in hit fraction 
    indicates that the duration of the test in cycles is beyond the fist step in cache latency as shown in 
    Figure~\ref{fig:cache-miss} (approx. 200 cycles). Whether or not the instruction sequence reaches the 
    cycle count limit we see only a sharp drop in hit fraction to 0 indicating the limiting factor 
    is ROB size. For example, the \texttt{idiv (r32)} instruction we identify 18 as the maximum number of 
    instruction limited by ROB (not including ~15 instructions for signal gadget), with a soft limit at 
    12 instructions enforced by the cache latency. We measure the ROB size by timing two cache 
    accesses with a variable number of instructions between them and note that this aligns with 
    the measured maximum \texttt{nop} instrustions.
    }
    \label{fig:spec-capacity-np}
\end{figure*}
}
